Utility of Diffusion Models for Predicting and Diagnosing the Spread of Conservation Behaviors

### 2021-11-23 DOI_Review/Manuscript.rmd
catching up for PR

date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Community-based conservation, diffusion of innovations, cultural evolution, complex systems
bibliography: [./DOI_Conservation_Review.bib]
abstract: A common goal of conservation practitioners is to design interventions that benefit both people and the environment, and as a result are adopted robustly by local communities. Diffusion theory provides generalized heuristics and models for how learned behaviors spread within a population and has been increasingly identified by researchers as useful for understanding conservation adoption. Here, we elaborate on the need for better understanding the spread of conservation behaviors, and review diffusion theory and its past application to these problems. We follow this with a practical demonstration of how the mathematical models which fit the diffusion process can be used to diagnose and predict the uptake of conservation behaviors even under tightly coupled behavior-environment feedbacks. Data on conservation adoption are generated through an agent-based simulation which highlights our dynamics of interest. Throughout this review, situate diffusion theory in the broader literature on cultural evolution and use these insights to refine our claims about how diffusion can be productively applied to conservation. Specifically, we show that the social learning mechanism individuals employ can alter the population pattern of adoption and the final proportion of individuals that adopt an intervention. Overally, we provide a quantitative approach for modeling the spread of conservation behaviors and argue for a more complete incorporation of cultural evolutionary theory when considering diffusion in these systems.

@@

When modifying their beliefs or behaviors, people do not imitate any other individual at random [@bandura_social_1977]. Instead, they preferentially adopt traits by employing a variety of learning biases such as success-biased learning where individuals who are perceived as more successful are more likely to be copied than unsuccessful individuals [@boyd_culture_1988; @laland_social_2004]. As a result, the cumulative culture of a population evolves over time as various beliefs and behaviors are unequally copied and their frequency within a population changes. While a full review of cultural evolution is beyond the scope of this paper, the general intuition is that populations carry a variable pool of information which is shaped by events in the lives of individuals such as social observation or natural selection [@mesoudi_cultural_2016]. This information affects individuals’ behaviors, which in-turn may affect their environments, and further alter which traits are adopted in the future [@durham_coevolution_1991]. Thus, cultural evolution provides an intuitive way to scale from individual learning and decision making to population level patterns and even population level behavior-environment feedbacks.

Diffusion of novel beliefs and behaviors is a specific instance of cultural evolution, where the trait being transmitted is new to the population and outcompetes some previous cultural variant, such as in the classic example of hybrid corn seed adoption in Iowa farmers [@mesoudi_cultural_2016; @ryan_acceptance_1950]. We observe these patterns through diffusion curves which show the proportion of a population that has adopted a belief or behavior on the Y-axis, as a function of time on the X-axis [@hoppitt_effect_2010; @rogers_diffusion_2010; @ryan_acceptance_1950]. The exact shape of these diffusion curves is highly dependent on the learning mechanism employed by individuals in the population, as well as contextual factors like the population density or ease of trialing the cultural variant [@smaldino_adoption_2017; @henrich_cultural_2001; @rogers_diffusion_2010 ].

In the vast majority of empirical studies, the diffusion of a novel trait produces a sigmoidal “S” shape, where it initially spreads slowly, followed by a period of rapid adoption, then again spreads slowly until it reaches fixation [@rogers_diffusion_2010 ]. This “S” shaped curve is indicative of biased social learning, in contrast to the relatively rare “R” shaped curves observed when individuals learn through independent trial and error (fig 1) [@henrich_cultural_2001]. These two curve shapes however are not completely discrete, as the combination of learning strategies employed by individuals in a population shifts more toward individual or social learning, we expect to see a shift toward an “R” shape or “S” shape respectively [@henrich_cultural_2001].

@@

# Diffusion Theory in Community-Based Conservation

Just as the sigmoidal pattern indicative of social learning is seen in the majority of cases on behavioral adoption broadly, scholars have also noticed this pattern in over 80% of case studies on the uptake of community-based conservation interventions worldwide [@mascia_when_2018; @mills_how_2019]. High-level studies have begun identifying diffusion theory as potentially useful for diagnosing why some community-based conservation interventions spread and persist, while others fizzle out. For example, in a review of 22 conservation interventions referenced above, the majority of initiatives were adopted by less than 30% of potential adopters. The authors point to a sigmoidal diffusion pattern to explain the slow start of many initiatives and suggest that practitioners may actively spread information on interventions (as in Abernethy et al. [-@abernethy_two_2014]), or facilitate communication to increase adoption rates [@mills_how_2019; @greenhalgh_diffusion_2004]. Further, a recent framework for understanding the establishment, persistence, and spread of community‐based conservation practices identifies qualitative insights, such as intervention observability, from diffusion theory as useful in guiding the development of conservation interventions [@mahajan_theorybased_2020].

@@

These studies are compelling in their conclusion that diffusion theory can be productively applied to the spread of community-based conservation, but have been critiqued as qualitative suggestions may miss important contextual factors unique to each intervention [@steenbergen_theory_2021-1; @lam_scaling_2020]. Below, we make a quantitative argument for using diffusion theory to understand the spread of community-based conservation interventions. We envision a practitioner who is monitoring the weekly, monthly, or even yearly uptake rates of an intervention and is hoping to predict if and when it will “take off” in the population, what the final proportion of adopters will be, and/or what they can do to promote its spread. The aim of this review is to explicitly demonstrate how diffusion theory can be used to generate predictions about the patterns of uptake of conservation behaviors given realistic data, and diagnose when in the lifetime of an intervention individuals may be dropping out.

# Quantitative applications

The sigmoidal pattern of behavioral spread over time mirrors that of disease transmission [@smaldino_coupled_2020]. This parallel implies that the well-established mathematics used to describe patterns of disease transmission can be applied to fitting models of behavioral spread [@cavalli-sforza_cultural_1981; @strang_adding_1991]. Specifically, the canonical susceptible-infectious-recovered (SIR) model and SIRS variation where individuals are added back to the pool of susceptibles after recovery (eq 1), have been demonstrated to be well suited to fitting the diffusion process [@famil-rohani_development_2019; @smaldino_coupled_2020]. These compartmental differential equations assign every member of a population to a category (S, I, or R), depending on the proportion of individuals in the other categories and parameter values for the natural history characteristics of a given disease [@tolles_modeling_2020]. In this instantiation, the natural history parameter $\beta$ represents the rate of infection in the population, $\gamma$ represents the rate of recovery, and $\chi$ is the rate that recovered individuals are added back into the pool of susceptibles.

\begin{align}
&\frac{dS}{dt} = - \beta \cdot \text{S}_{t} \cdot \text{I}_{t} + \chi \cdot \text{R}_{t} \nonumber\\
&\frac{dI}{dt} = \beta \cdot \text{S}_{t} \cdot \text{I}_{t} - \gamma \cdot \text{I}_{t} \\
&\frac{dR}{dt} = \gamma \cdot \text{I}_{t} - \chi \cdot \text{R}_{t} \nonumber
\end{align}

When the natural history characteristics of some pathogen is known, the SI, SIR, SIS, & SIRS suite of models reliably give accurate, deterministic predictions about the number of individuals in a population who will become infected and for how long a disease will persist within a population [@huppert_mathematical_2013]. For conservation interventions however, there are no natural history features of a specific intervention, such as the rate or duration of adoption, which can be known ahead of time. We can however, use Bayesian inference to probabilistically estimate these parameters from data on behavioral uptake over time [@margossian_review_2019; @margossian_differential_2017; @carpenter_predator-prey_2018].

Below, we develop a minimal agent-based model to highlight our core dynamics of interest. In this abstraction, we simulate the spread in adoption of a conservation intervention from just a few initial individuals, to a stable proportion of the population. We use this simulation to produce realistic data on long-term adoption dynamics under tightly coupled behavior-environment feedbacks unique to community-based conservation interventions. Then, using just the first 30% of the data, we probabilistically estimate the rate of intervention spread, rate of dropout, and readoption using the SIRS model shown above. We use these estimated parameters to predict the long-term patterns of adoption for the remaining 70% of time steps and compare our predicted values to those observed in the simulation.

# Agent-based simulation
We developed a minimal agent-based simulation which we use to generate data on conservation adoption within a population. We use these data to test the performance of the SIRS diffusion model in predicting patterns of conservation behavior adoption under two social learning mechanisms. The first is unbiased copying, where agents adopt an alternative behavior with a probability proportional to its frequency in the population, independent of the benefit provided by adopting the behavior or not [@barrett_equifinality_2019]. The second mechanism is success-biased learning, under which agents disproportionately copy the behaviors of more successful agents [@baldini_two_2013]. This mechanism is formally demonstrated in equation 1, for two behaviors: $A$ & $B$, where the probability that an individual adopts behavior $A$ is a function of the mean pay-off, $\bar{P}$, of both behaviors at each time step and the strength of the bias, $\beta$ [@barrett_equifinality_2019].

\begin{equation}
 \text{Pr}(A|A,B)_{t+1} = \frac{\text{exp}\beta\bar{P}_{A,t}}{\text{exp}\beta\bar{P}_{A,t} + \text{exp}\beta\bar{P}_{B,t}}  
\end{equation}

Our simulated system emulates the common performance payments or payments for ecosystem services approach to conservation [@ferraro_future_2011]. We establish a degraded landscape/seascape (hereafter referred to as landscape) where all simulated individuals are dependent on harvesting some renewable, mobile resource. In the first time step, we introduce a conservation intervention which delineates 20% of the landscape as a conservation area and is initially adopted by just 2% of the population. In each subsequent time step, agents enrolled in the conservation program receive a reward, and continue to harvest just from the remaining 80% of the unconserved landscape. Agents not enrolled in the intervention harvest from the entire landscape and forgo the reward. After each harvest period, agents are randomly paired, and employ either unbiased or success-biased learning to decide whether or not to adopt the conservation behavior in the next time step [@baldini_success-biased_2012]. The conservation and unconserved resource pools are reduced according to agents’ harvesting behavior, regenerate, and move from one area to the other if the carrying capacity is reached in any respective area. As behavioral adoption under success-biased learning is dependent on harvest success, the success of each behavior is determined largely by the amount of available resources in each area, and the amount of resources in each area is a result of behavioral adoption, this learning strategy is contingent on behavior-environment feedbacks, unlike unbiased adoption which occurs independent of harvest success.

@@

A visual description of this minimal model is shown in figure 1, and the full ODD+ model description can be found in the supplemental material (S1). Interested readers can find an interactive web application which allows users to run this simulation given a wide variety of starting conditions at [https://matthewclark.shinyapps.io/ConservationAdoptionSimulation/](https://matthewclark.shinyapps.io/ConservationAdoptionSimulation/). As this simulation is not the emphasis of this study, we show just two runs below. The starting conditions of these runs are identical except for the social learning mechanism employed by the population. As the unbiased learning mechanism is independent of the resource levels, the adoption curve follows the traditional “S” pattern and nears fixation in the long-term. Under environmentally mediated success-biased learning, the adoption pattern spikes initially, but drops off dramatically as the resources in the unconserved landscape are reduced following increased harvest pressure there as a result of the conservation intervention. The full output of this latter simulation is shown below in figure 2.

>>>>

date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Community-based conservation, diffusion of innovations, cultural evolution, complex systems
bibliography: [./DOI_Conservation_Review.bib]
abstract: A common goal of conservation practitioners is to design interventions that benefit both people and the environment, and as a result are adopted robustly by local communities. Diffusion theory provides generalized heuristics and models for how learned behaviors spread within a population and has been increasingly identified by researchers as useful for understanding conservation adoption. Here, we elaborate on the need for better understanding the spread of conservation behaviors and review diffusion theory and its past application to these problems. We follow this with a practical demonstration of how the mathematical models which fit the diffusion process can be used to diagnose and predict the uptake of conservation behaviors even under tightly coupled behavior-environment feedbacks. Data on conservation adoption are generated through an agent-based simulation which highlights our dynamics of interest. Throughout this review, situate diffusion theory in the broader literature on cultural evolution and use these insights to refine our claims about how diffusion can be productively applied to conservation. Specifically, we show that the social learning mechanism individuals employ can alter the population pattern of adoption and the final proportion of individuals that adopt an intervention. Overally, we provide a quantitative approach for modeling the spread of conservation behaviors and argue for a more complete incorporation of cultural evolutionary theory when considering diffusion of community-based conservation initiatives.

@@

When modifying their beliefs or behaviors, people do not imitate any other individuals at random [@bandura_social_1977]. Instead, they preferentially adopt traits by employing a variety of learning biases such as success-biased learning where individuals who are perceived as more successful are more likely to be copied than unsuccessful individuals [@boyd_culture_1988; @laland_social_2004]. As a result, the cumulative culture of a population evolves over time as various beliefs and behaviors are unequally copied and their frequency within a population changes. While a full review of cultural evolution is beyond the scope of this paper, the general intuition is that populations carry a variable pool of information which is shaped by events in the lives of individuals such as social observation or natural selection [@mesoudi_cultural_2016]. This information affects individuals’ behaviors, which in-turn may affect their environments, and further alter which traits are adopted in the future [@durham_coevolution_1991]. Thus, cultural evolution provides an intuitive way to scale from individual learning and decision making to population level patterns and even population level behavior-environment feedbacks.

Diffusion of novel beliefs and behaviors is a specific instance of cultural evolution, where the trait being transmitted is new to the population and outcompetes some previous cultural variant, such as in the classic example of hybrid corn seed adoption in Iowa farmers [@mesoudi_cultural_2016; @ryan_acceptance_1950]. We observe these patterns through diffusion curves which show the proportion of a population that has adopted a belief or behavior on the Y-axis, as a function of time on the X-axis [@hoppitt_effect_2010; @rogers_diffusion_2010; @ryan_acceptance_1950]. The exact shape of these diffusion curves is highly dependent on the learning mechanism employed by individuals in the population, as well as contextual factors like the population density or ease of trialing the cultural variant [@smaldino_adoption_2017; @henrich_cultural_2001; @rogers_diffusion_2010].

In the vast majority of empirical studies, the diffusion of a novel trait produces a sigmoidal “S” shape, where it initially spreads slowly, followed by a period of rapid adoption, then again spreads slowly until it reaches fixation [@rogers_diffusion_2010]. This “S” shaped curve is indicative of biased social learning, in contrast to the relatively rare “R” shaped curves observed when individuals learn through independent trial and error (fig 1) [@henrich_cultural_2001]. These two curve shapes however are not completely discrete, as the combination of learning strategies employed by individuals in a population shifts more toward individual or social learning, we expect to see a shift toward an “R” shape or “S” shape respectively [@henrich_cultural_2001].

@@

# Diffusion Theory in Community-Based Conservation
Just as the sigmoidal pattern indicative of social learning is seen in the majority of cases on behavioral adoption broadly, scholars have also noticed this pattern in over 80% of case studies on the uptake of community-based conservation interventions worldwide [@mascia_when_2018; @mills_how_2019]. High-level studies have begun identifying diffusion theory as potentially useful for diagnosing why some community-based conservation interventions spread and persist, while others fizzle out. For example, in a review of 22 conservation interventions referenced above, the majority of initiatives were adopted by less than 30% of potential adopters. The authors point to a sigmoidal diffusion pattern to explain the slow start of many initiatives and suggest that practitioners may actively spread information on interventions (as in Abernethy et al. [-@abernethy_two_2014]), or facilitate communication to increase adoption rates [@mills_how_2019; @greenhalgh_diffusion_2004]. Further, a recent framework for understanding the establishment, persistence, and spread of community‐based conservation practices identifies qualitative insights, such as intervention observability, from diffusion theory as useful in guiding the development of conservation interventions [@mahajan_theorybased_2020].

@@

These studies are compelling in their conclusion that diffusion theory can be productively applied to the spread of community-based conservation, but have been critiqued as qualitative suggestions may miss important contextual factors unique to each intervention [@steenbergen_theory_2021-1; @lam_scaling_2020]. Below, we make a quantitative argument for using diffusion theory to understand the spread of community-based conservation interventions. We envision a practitioner who is monitoring the weekly, monthly, or even yearly uptake rates of an intervention and is hoping to predict if and when it will “take off” in the population, what the final proportion of adopters will be, and what they can do to promote its spread. The aim of this review is to explicitly demonstrate how diffusion theory can be used to generate predictions about the patterns of uptake of conservation behaviors given realistic data and diagnose when in the lifetime of an intervention individuals may be dropping out.

# Quantitative applications
In this section we explain how we can fit the pattern of diffusion mathematically to derive inference about the rate and degree to which an intervention will be adopted by a population. The “S” shaped diffusion pattern of behavioral spread over time mirrors that of community disease transmission [@smaldino_coupled_2020]. This parallel implies that the well-established mathematics used to describe patterns of disease transmission can be applied to fitting models of behavioral spread [@cavalli-sforza_cultural_1981; @strang_adding_1991]. Specifically, the canonical susceptible-infectious-recovered (SIR) model and SIRS variation where individuals are added back to the pool of susceptibles after recovery (eq 1), have been demonstrated to be well suited to fitting the diffusion process [@famil-rohani_development_2019; @smaldino_coupled_2020]. These compartmental differential equations assign every member of a population to a category (S, I, or R), depending on the proportion of individuals in the other categories and parameter values for the natural history characteristics of a given disease [@tolles_modeling_2020]. In this instantiation, the natural history parameter $\beta$ represents the rate of infection in the population, $\gamma$ represents the rate of recovery, and $\chi$ is the rate that recovered individuals are added back into the pool of susceptibles.

\begin{align*}
&\frac{dS}{dt} = - \beta \cdot \text{S}_{t} \cdot \text{I}_{t} + \chi \cdot \text{R}_{t} \\
&\frac{dI}{dt} = \beta \cdot \text{S}_{t} \cdot \text{I}_{t} - \gamma \cdot \text{I}_{t} \\
&\frac{dR}{dt} = \gamma\cdot \text{I}_{t} - \chi \cdot \text{R}_{t}
\end{align*}

When the natural history characteristics of some pathogen are known, the SI, SIR, SIS, & SIRS suite of models reliably give accurate, deterministic predictions about the number of individuals in a population who will become infected and for how long a disease will persist within a population [@huppert_mathematical_2013]. For conservation interventions however, there are no natural history features of a specific intervention, such as the rate or duration of adoption, which can be known ahead of time. Instead, the qualitative insights from diffusion theory referenced above aim to increase the rate of spread and reduce the rate of dropout so that the intervention persists in the population and may even spread to new communities. While we can not know exactly what the rates of spread and dropout will be ahead of time, we can use Bayesian inference to probabilistically estimate them once the project has been implemented [@margossian_review_2019; @margossian_differential_2017; @carpenter_predator-prey_2018]. Knowing these parameters may allow us to predict the final rate of uptake from early in the project, identify whether adoption or dropout are the primary issue, and apply the qualitative insights from diffusion theory more targetedly.

In some contexts, the uptake of community-based conservation is subject to behavior-environment feedbacks, a fundamental difference from behavioral adoption in the abstract [@liu_complexity_2007]. In the case of marine protected areas for example, success in conserving fish populations may actually disincentivize individuals to continue supporting conservation programs [@christie_marine_2004]. A key question then is whether the diffusion models explained above will still reliably fit the spread of conservation behaviors when those behaviors alter the environment, and changes in the environment alter the payoff structure of various behavioral choices. To answer this question, we use an agent-based simulation to generate data on the spread of a conservation behavior via social learning, contingent on feedbacks with the environment. Then, using just the first 30% of the data, we probabilistically estimate the rate of intervention spread, rate of dropout, and readoption using a Bayesian instantiation of the SIRS model shown above. We use these estimated parameters to predict the long-term patterns of adoption for the remaining 70% of time steps and compare our predicted values to those observed in the simulation. While the purpose of this example is to test whether the SIRS model can capture our core dynamics of interest, it also serves as an explicit demonstration of how a practitioner or other researchers can directly apply the quantitative process described above. We first briefly describe the simulation below.

# Agent-based simulation
We developed a minimal agent-based simulation which we use to generate data on conservation adoption within a population. We use these data to test the performance of the SIRS diffusion model in fitting patterns of conservation behavior adoption under two social learning mechanisms. The first is unbiased, frequency-based, copying where agents adopt an alternative behavior with a probability proportional to its frequency in the population, independent of the benefit provided by adopting the behavior or not [@barrett_equifinality_2019]. The second mechanism is success-biased learning, under which agents disproportionately copy the behaviors of more successful agents [@baldini_two_2013]. This mechanism is formally demonstrated in equation 1, for two behaviors: $A$ & $B$, where the probability that an individual adopts behavior $A$ is a function of the mean pay-off, $\bar{P}$, of both behaviors at each time step and the strength of the bias, $\beta$ [@barrett_equifinality_2019].

$$ \text{Pr}(A|A,B)_{t+1} = \frac{\text{exp}\beta\bar{P}_{A,t}}{\text{exp}\beta\bar{P}_{A,t} + \text{exp}\beta\bar{P}_{B,t}} $$

Our simulated system emulates the common performance payments or payments for ecosystem services approach to conservation [@ferraro_future_2011; @rakotonarivo_role_2021]. We establish a degraded landscape/seascape (hereafter referred to as landscape) where all simulated individuals are dependent on harvesting some renewable, mobile resource. In the first time step, we introduce a conservation intervention which delineates 20% of the landscape as a conservation area and is initially adopted by just 2% of the population. In each subsequent time step, agents enrolled in the conservation program receive a reward, and continue to harvest just from the remaining 80% of the unconserved landscape. Agents not enrolled in the intervention harvest from the entire landscape and forgo the reward. After each harvest period, agents are randomly paired, and employ either unbiased or success-biased learning to decide whether or not to adopt the conservation behavior of their partner in the next time step [@baldini_success-biased_2012]. The conserved and unconserved resource pools are reduced according to agents’ harvesting behavior, regenerate, and move from one area to the other if the carrying capacity is reached in any respective area. As behavioral adoption under success-biased learning is dependent on harvest success, the success of each behavior is determined largely by the amount of available resources in each area, and the amount of resources in each area is a result of behavioral adoption, this learning strategy is contingent on behavior-environment feedbacks, unlike unbiased adoption which occurs independent of harvest success.

@@

A visual description of this minimal model is shown in figure 2, and the full ODD+ model description can be found in the supplemental material (S1). Interested readers can find an interactive web application which allows users to run this simulation given a wide variety of starting conditions at [https://matthewclark.shinyapps.io/ConservationAdoptionSimulation/](https://matthewclark.shinyapps.io/ConservationAdoptionSimulation/). As this simulation is not the emphasis of this study, we show just two runs below. The starting conditions of these runs are identical except for the social learning mechanism employed by the population. As the unbiased learning mechanism is independent of the resource levels, the adoption curve follows the traditional “S” pattern and nears fixation in the long-term. Under environmentally mediated success-biased learning, the adoption pattern spikes initially, but drops off dramatically as the resources in the unconserved landscape are reduced following increased harvest pressure there as a result of the conservation intervention. The full output of this latter simulation is shown below in figure 3.
